---
title: "Student Performance Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Student Performance Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/zaraavagyan/weathercsv/discussion\>*

### Reference:

*\<Avagyan, Z. (2017). Weather CSV [Data set]. Kaggle. https://www.kaggle.com/datasets/zaraavagyan/weathercsv\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))
## Loading the Dataset
```{r Load dataset}
# Load dataset
StudentData <- read.csv("student_performance.csv", colClasses = c(
  Gender = "factor",
  AttendanceRate = "numeric",
  StudyHoursPerWeek = "numeric",
  PreviousGrade = "numeric",
  ExtracurricularActivities = "numeric",
  ParentalSupport = "factor",
  FinalGrade = "numeric"
))
# Preview structure and data
str(StudentData)
head(StudentData)
View(StudentData)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency of Gender
table(StudentData$Gender)

# Frequency of Parental Support levels
table(StudentData$ParentalSupport)

```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Central Tendency: Mean & Median
mean(StudentData$AttendanceRate)
median(StudentData$AttendanceRate)

mean(StudentData$StudyHoursPerWeek)
median(StudentData$StudyHoursPerWeek)

mean(StudentData$FinalGrade)
median(StudentData$FinalGrade)

# Mode function (for categorical/factor or numeric variables)
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

get_mode(StudentData$ParentalSupport)
get_mode(StudentData$Gender)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Standard Deviation
sd(StudentData$AttendanceRate)
sd(StudentData$StudyHoursPerWeek)
sd(StudentData$FinalGrade)

# Range
range(StudentData$AttendanceRate)
range(StudentData$StudyHoursPerWeek)
range(StudentData$FinalGrade)

# IQR (Interquartile Range)
IQR(StudentData$AttendanceRate)
IQR(StudentData$StudyHoursPerWeek)
IQR(StudentData$FinalGrade)

```

## Measures of Relationship
```{r MOR}
# Correlation matrix for numeric predictors and FinalGrade
numeric_vars <- StudentData[, sapply(StudentData, is.numeric)]
cor(numeric_vars)
```

## ANOVA
```{r ANOVA}
# Visual correlation plot (optional)
# install.packages("corrplot") if needed
library(corrplot)
corrplot(cor(numeric_vars), method = "number")

# ANOVA to check if FinalGrade differs by Gender
anova_gender <- aov(FinalGrade ~ Gender, data = StudentData)
summary(anova_gender)

# ANOVA to check if FinalGrade differs by Parental Support
anova_parental_support <- aov(FinalGrade ~ ParentalSupport, data = StudentData)
summary(anova_parental_support)

# Post-hoc Tukey's HSD for ParentalSupport
tukey_result <- TukeyHSD(anova_parental_support)
summary(tukey_result)

# Post-hoc Tukey's HSD for Gender (optional, if needed)
tukey_gender <- TukeyHSD(anova_gender)
summary(tukey_gender)
```

## Plots
```{r Plots}
# Load ggplot2 for plotting
library(ggplot2)

# Histogram for FinalGrade (continuous variable)
ggplot(StudentData, aes(x = FinalGrade)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of FinalGrade", x = "FinalGrade", y = "Frequency")

# Boxplot for StudyHoursPerWeek (continuous variable)
ggplot(StudentData, aes(x = "", y = StudyHoursPerWeek)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Boxplot of StudyHoursPerWeek", y = "Study Hours Per Week")

# Bar plot for Gender (categorical variable)
ggplot(StudentData, aes(x = Gender)) +
  geom_bar(fill = "lightgreen", color = "black") +
  labs(title = "Bar Plot of Gender", x = "Gender", y = "Count")

# Bar plot for ParentalSupport (categorical variable)
ggplot(StudentData, aes(x = ParentalSupport)) +
  geom_bar(fill = "lightcoral", color = "black") +
  labs(title = "Bar Plot of Parental Support", x = "Parental Support", y = "Count")

# Scatter plot for StudyHoursPerWeek vs FinalGrade (continuous vs continuous)
ggplot(StudentData, aes(x = StudyHoursPerWeek, y = FinalGrade)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of Study Hours vs Final Grade", x = "Study Hours Per Week", y = "Final Grade") +
  geom_smooth(method = "lm", color = "red")  # Adds a linear regression line

# Scatter plot faceted by Gender
ggplot(StudentData, aes(x = StudyHoursPerWeek, y = FinalGrade)) +
  geom_point(aes(color = Gender)) +
  labs(title = "Study Hours vs Final Grade by Gender", x = "Study Hours Per Week", y = "Final Grade") +
  facet_wrap(~ Gender)  # Facets the plot by Gender

# Boxplot for FinalGrade by ParentalSupport
ggplot(StudentData, aes(x = ParentalSupport, y = FinalGrade, fill = ParentalSupport)) +
  geom_boxplot() +
  labs(title = "Boxplot of FinalGrade by Parental Support", x = "Parental Support", y = "Final Grade") +
  scale_fill_brewer(palette = "Set3")

# Load the corrplot package
library(corrplot)

# Select continuous variables for correlation
cont_vars <- StudentData[, c("AttendanceRate", "StudyHoursPerWeek", "PreviousGrade", "FinalGrade")]

# Compute the correlation matrix
cor_matrix <- cor(cont_vars)

# Plot the correlation matrix
corrplot(cor_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

```

# Preprocessing & Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in the entire dataset
missing_values <- colSums(is.na(StudentData))

# Print the number of missing values in each column
print(missing_values)

# If you want to see if there are any columns with missing values
if (any(missing_values > 0)) {
  cat("There are missing values in the following columns:\n")
  print(names(missing_values[missing_values > 0]))
} else {
  cat("There are no missing values in the dataset.")
}

# Load VIM package
library(VIM)

# Visualize missing data
aggr_plot <- aggr(StudentData, col=c('skyblue', 'red'), numbers=TRUE, sortVars=TRUE, labels=names(StudentData), cex.axis=0.7, gap=3, ylab=c("Missing data", "Pattern"))

```

# Training Model
## LOOCV
```{r LOOCV}
# Load necessary libraries
library(caret)

# Set up LOOCV using caret's trainControl
ctrl <- trainControl(method = "LOOCV")
```

## Train different Models
```{r Train different Models}
# Train a Random Forest model using LOOCV
rf_model <- train(FinalGrade ~ ., 
                  data = StudentData, 
                  method = "rf", 
                  trControl = ctrl)

# Print model results
print(rf_model)

# Train a K-Nearest Neighbors model using LOOCV
knn_model <- train(FinalGrade ~ ., 
                   data = StudentData, 
                   method = "knn", 
                   trControl = ctrl)

# Print results for KNN
print(knn_model)
```

## Performance Comparison
```{r Performance Comparison}
# Load necessary library
library(ggplot2)

# Extract model performance
rf_results <- rf_model$results
knn_results <- knn_model$results

# Select the best tuning parameter results
rf_best <- rf_results[which.min(rf_results$RMSE), ]
knn_best <- knn_results[which.min(knn_results$RMSE), ]

# Create a dataframe for plotting
performance_df <- data.frame(
  Model = c("RandomForest", "KNN"),
  RMSE = c(rf_best$RMSE, knn_best$RMSE),
  Rsquared = c(rf_best$Rsquared, knn_best$Rsquared)
)

# Plot RMSE comparison
ggplot(performance_df, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "RMSE Comparison", y = "Root Mean Squared Error") +
  theme_minimal()

# Plot R-squared comparison
ggplot(performance_df, aes(x = Model, y = Rsquared, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "R-squared Comparison", y = "R²") +
  theme_minimal()
```

## Saving Model
```{r Saving Model}
# Saving the Random Forest model
saveRDS(rf_model, "./models/saved_rf_model_student.rds")

# Load the saved model
loaded_rf_model_student <- readRDS("./models/saved_rf_model_student.rds")

# Model predicts FinalGrade
new_student <- data.frame(
  Gender = factor("Male", levels = levels(StudentData$Gender)),
  AttendanceRate = 85,
  StudyHoursPerWeek = 15,
  PreviousGrade = 78,
  ExtracurricularActivities = 2,
  ParentalSupport = factor("High", levels = levels(StudentData$ParentalSupport))
)

# Use the loaded model to make predictions
predicted_grade <- predict(loaded_rf_model_student, newdata = new_student)

# Print prediction
print(predicted_grade)

```

